---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dplyover

<!-- badges: start -->
[![CodeFactor](https://www.codefactor.io/repository/github/timteafan/dplyover/badge)](https://www.codefactor.io/repository/github/timteafan/dplyover)
[![Codecov test coverage](https://codecov.io/gh/TimTeaFan/dplyover/branch/main/graph/badge.svg)](https://codecov.io/gh/TimTeaFan/dplyover?branch=main)
<!-- badges: end -->


## Overview
{dplyover} extends {dplyr}'s functionality by building a function family
around `dplyr::across()`.

The goal of this <over-across function family> is to provide a concise and uniform
syntax which can be used to create columns by applying functions to vectors and / or
sets of columns in {dplyr}. Ideally, this will:

- *improve our mental model* so that it is easier to tackle problems where the
solution is based on creating new columns.
- *reduce the amount of code* to create variables derived from existing colums 
when doing explanatory data analysis (EDA) (e.g. lagging, collapsing many
variables in a similar way). 
- *provide a clean {dplyr} approach* to create many variables which are calculated
based on two or more varialbes. 

The functions in the <over-apply function family> create columns by applying
one or several functions to:

### basic functions
 - `dplyr::across()`: a set of columns (not part of {dplyover})
 - `over()`: a vector (list or atomic vector)

### variants
 - `over2()` two vectors of the same length (pairwise)
 - `over2x()` two vectors (nested)
 - `across2()` two sets of columns (pairwise)
 - `across2x()` two sets of columns (nested)
 - `crossover()` a set of columns and a vector (pairwise)
 - `crossoverx()` a set of columns and a vector (nested)


## Installation

{dplyover} is not on CRAN. You can install the latest version from 
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("TimTeaFan/dplyover")
```

## Getting started

Below are a few examples of the <over-across function family>. More functions
and workarounds of how to tackle the problems below without {dplyover} can be
found in the vignette "Why dplyover?".

```{r, setup}
# dplyover is extention of dplyr:
library(dplyr)
library(dplyover)

# For better printing:
iris <- as_tibble(iris)
```

#### Applying a function with a varying argument to one column

`over()` applies one or several functions to a vector. We can use it inside
`dplyr::mutate()` to create several similar variables that we derive from an
existing column. This is helpful in cases where we want to create a batch of
similar variables with only slightly changes in the arguments of the calling
function. A good example are `lag` and `lead` variables. Below use column 'a'
to create lag and lead variables by `1`, `2` and `3` positions. `over()`'s 
`.names` argument lets us put nice names on the output columns.

```{r} 
tibble(a = 1:10) %>%
  mutate(over(c(1:3),
              list(lag  = ~ lag(a, .x),
                   lead = ~ lead(a, .x)),
              .names = "a_{fn}{x}"))
```

#### Applying a function with a varying argument to a set of columns

`crossoverx()` applies the functions in `.fns` to every combination of colums in
`.xcols` and elements in `.y`. This is similar to the example above, but this time,
we use a set of columns. Below we create five lagged variables for each
'Sepal.Length' and 'Sepal.Width'. To create nice names we use a named list as
argument in `.fns` and specify the glue syntax in `.names.`

```{r}
iris %>%
   transmute(
     crossoverx(starts_with("sepal"),
                1:5,
                list(lag = ~ lag(.x, .y)),
                .names = "{xcol}_{fn}{y}")) %>%
   glimpse
```


#### Applying functions to a pair of variables

`across2()` can be used to transform pairs of variables in one or more functions.
In the example below we want to calculate the product and the sum of all pairs
of 'Length' and 'Width' variables. We can use `{pre}` in the glue specification
in `.names` to extract the common prefix of each pair of variables. We can further
transform the names, in the example setting them `tolower` by specifying the
`.names_fn` argument:

```{r}
iris %>%
  transmute(across2(ends_with("Length"),
                    ends_with("Width"),
                    .fns = list(product = ~ .x * .y,
                                sum = ~ .x + .y),
                   .names = "{pre}_{fn}",
                   .names_fn = tolower))
```


## Performance and Compability

This is an experimental package which I started developing with my own use cases
in mind. I tried to keep the effort low, which is why this package *does not* 
internalize (read: copy) internal dplyr functions (especially the 'context internals').
This made it relatively easy to develop the package without copying tons of dplyr
code and overwritting existing functions, such as `mutate` and other one table
verbs.

However, the downside is that not relying on {dplyr} internals has some negative
effects in terms of performance and compability.

In a nutshell this means:

 - the over-across function family in {dplyover} is somewhat slower than the
 original `dplyr::across`.
 - Although {dplyover} is designed to work (only) in {dplyr}, some features and
 edge cases will not work correctly.
  
The good news is, that even without relying on {dplyr} internals most of the
original functionality can be replicated, and although being a bit less performant,
the current setup is optimized to fall not too far behind in terms of speed.

Regarding compability, I have spent a serious amount of time on testing by
replicating most of the tests for `dplyr::across` successfully. 

For more information on the performance and compability of {dplyover} see the
vignette "Performance and Compability".


## History

I originally opened a
[feature request on GitHub](https://github.com/tidyverse/dplyr/issues/4834) to
include a very special case version of `over` (or to that time `mutate_over`)
into {dplyr}. The adivse then was to make this kind of functionality available
in a separate package. While I was working on this very special case version of
`over` I realized that the general use case looks more like a `purrr::map` function
for inside {dplyr} verbs with different variants, which led me to the over-across
function family.


## Acknowledgements and Disclaimer

This package is not only an extention of {dplyr}. The main functions in
{dplyover} are directly derived and based on `dplyr::across()` (dplyr's license
and copyrights apply). So if this package is working correctly, all the credit
should go to the dplyr team. 

My own contribution merely consists of: 

 1. removing the underlying dependencies on dplyr's internal functions, and
 2. slightly changing `across`' logic to make it work for vectors and or a
combination of two vectors and / or sets columns.

By this I probably managed to introduce some bugs and edge cases which won't work, 
and in which case I am the only one to blame.
