---
title: "Performance and Compability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance and Compability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bench}
  %\VignetteDepends{ggplot2}
  %\VignetteEncoding{UTF-8}
---

```{r, setup}
options(
  tibble.print_min = 4,
  tibble.max_extra_cols = 8,
  digits = 2,
  crayon.enabled = FALSE,
  cli.unicode = FALSE
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidyr)
library(dplyover)
library(ggplot2)
library(bench)
```

# Introduction

Although {dplyover} is an extension of {dplyr}, it doesn't make use of (read: copy)
{dplyr} internals. This made it relatively easy to develop the package without
copying tons of dplyr code and overwritting existing functions, such as `mutate`
and other one table verbs. However, this has some negative side effects in terms
of performance and compability.

In a nutshell this means:
  - the over-across function family in {dplyover} is somewhat slower than the 
  original `dplyr::across`.
  - Although {dplyover} is designed to work (only) in {dplyr}, some features and
  edge cases will not work correctly.
  
The good news is, that even without relying on {dplyr} internals most of the
original functionality can be replicated, and although being a bit less performant,

... it is still doing ok compared to more verbose alternative approaches.


# Compability

 - cur_column()
 - two function calls inside the same expression
 - 

# Performance

{dplyover} is build around `dplyr::across()`. Therefore, we should first look at 
the original `across` function and how it compares to alternative approaches.

We take the `diamonds` data set from the {ggplot2} package and apply a "typical"
`across` transformation: we calculate the mean of three different variables. We
can do this either in `dplyr::across` or 'manually' by adding three new variables
inside a `dplyr::mutate` call.

```{r, eval = FALSE}
diamonds %>%
  summarise(across(c(x,y,z),
                   mean))
diamonds %>% 
  summarise(x = mean(x),
            y = mean(y),
            z = mean(z))
```

If we compare the performance of both operations, we can see that `across` has
a small overhead compared to the manual operation. This is due to `across` internal 
setup which contains a couple of checks and the column names setup.

```{r, echo = FALSE, eval = TRUE}
bench::mark(iterations = 100, 
  "across" = {
    diamonds %>%
      summarise(across(c(x,y,z),
                       mean))}, 
  "manually" = {
    diamonds %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))}
)
```

While the difference is small when comparing **one** call, this overhead
accumulates when doing `group_by` operations. To demonstrate this, we slice
the top 1, 10, 100, 500 and 1000 rows off the `diamonds` data set, group by
row with `rowwise`, and compare the same operations again.

Below is a code snippet for the 100 groups / rows case:
```{r, eval = FALSE}

diamonds_100rows <- diamonds %>%
  slice_head(n = 100) %>% 
  rowwise()

diamonds_100rows %>%
  summarise(across(c(x,y,z),
                   mean))

diamonds_100rows %>% 
  summarise(x = mean(x),
                y = mean(y),
                z = mean(z))
```

The plot below shows the median time in miliseconds that each operation (`across`
and the 'manual' approach) takes by increasing group size.

```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}

diamonds_1row <- diamonds %>%
  slice_head(n = 1) %>% 
  rowwise()

diamonds_10rows <- diamonds %>%
  slice_head(n = 10) %>% 
  rowwise()

diamonds_100rows <- diamonds %>%
  slice_head(n = 100) %>% 
  rowwise()

diamonds_500rows <- diamonds %>%
  slice_head(n = 500) %>% 
  rowwise()

diamonds_1000rows <- diamonds %>%
  slice_head(n = 1000) %>% 
  rowwise()

across_bench <- bench::mark(iterations = 50,
            check = FALSE,
  "across 1" = {
   diamonds_1row %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 10" = {
   diamonds_10rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 100" = {
   diamonds_100rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 500" = {
   diamonds_500rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 1000" = {
   diamonds_1000rows %>%
     summarise(across(c(x,y,z),
                   mean))},
   "manually 1" = {
    diamonds_1row %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 10" = {
    diamonds_10rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 100" = {
    diamonds_100rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 500" = {
    diamonds_500rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 1000" = {
    diamonds_1000rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))
    })

across_bench %>%
  separate(expression,
           into = c("fun", "groups"),
           convert = TRUE) %>% 
  ggplot(aes(x = groups, y = median, group = fun, color = fun)) +
  geom_line() + 
  scale_y_bench_time(base = NULL)

```

- compare across and over and across2




```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}

diamonds_1row <- diamonds %>%
  slice_head(n = 1) %>% 
  rowwise()

diamonds_10rows <- diamonds %>%
  slice_head(n = 10) %>% 
  rowwise()

diamonds_100rows <- diamonds %>%
  slice_head(n = 100) %>% 
  rowwise()

diamonds_500rows <- diamonds %>%
  slice_head(n = 500) %>% 
  rowwise()

diamonds_1000rows <- diamonds %>%
  slice_head(n = 1000) %>% 
  rowwise()

over_bench <- bench::mark(iterations = 50, check = FALSE,
  "across 1" = {
    diamonds_1row %>%
      summarise(across(x, ~ 1, .names = "{col}_1"))}, 
  "across 10" = {
    diamonds_10rows %>%
      summarise(across(x, mean, .names = "mean_{col}"))}, 
  "across 100" = {
    diamonds_100rows %>%
      summarise(across(x, mean, .names = "mean_{col}"))}, 
  "across 500" = {
    diamonds_500rows %>%
      summarise(across(x, mean, .names = "mean_{col}"))}, 
  "across 1000" = {
    diamonds_1000rows %>%
      summarise(across(x, mean, .names = "mean_{col}"))}, 
  "over 1" = {
    diamonds_1row %>% 
      summarise(over("1", ~ mean(x), .names = "mean_x"))},
  "over 10" = {
    diamonds_10rows %>% 
      summarise(over("1", ~ mean(x), .names = "mean_x"))},
  "over 100" = {
    diamonds_100rows %>% 
      summarise(over("1", ~ mean(x), .names = "mean_x"))},
  "over 500" = {
    diamonds_500rows %>% 
      summarise(over("1", ~ mean(x), .names = "mean_x"))},
  "over 1000" = {
    diamonds_1000rows %>% 
      summarise(over("1", ~ mean(x), .names = "mean_x"))}
)

over_bench %>%
  separate(expression,
           into = c("fun", "groups"),
           convert = TRUE) %>% 
  ggplot(aes(x = groups, y = median, group = fun, color = fun)) +
  geom_line() + 
  scale_y_bench_time(base = NULL)
```


- compare simmilar operations
   over and map (lag)
   across/cur_column and across2 (add two variables)


```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}

diamonds_1row <- diamonds %>%
  slice_head(n = 1) %>% 
  rowwise()

diamonds_10rows <- diamonds %>%
  slice_head(n = 10) %>% 
  rowwise()

diamonds_100rows <- diamonds %>%
  slice_head(n = 100) %>% 
  rowwise()

diamonds_500rows <- diamonds %>%
  slice_head(n = 500) %>% 
  rowwise()

diamonds_1000rows <- diamonds %>%
  slice_head(n = 1000) %>% 
  rowwise()

across_bench <- bench::mark(iterations = 50,
            check = FALSE,
  "across 1" = {
   diamonds_1row %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 10" = {
   diamonds_10rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 100" = {
   diamonds_100rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 500" = {
   diamonds_500rows %>%
     summarise(across(c(x,y,z),
                   mean))}, 
  "across 1000" = {
   diamonds_1000rows %>%
     summarise(across(c(x,y,z),
                   mean))},
   "manually 1" = {
    diamonds_1row %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 10" = {
    diamonds_10rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 100" = {
    diamonds_100rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 500" = {
    diamonds_500rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))},
  "manually 1000" = {
    diamonds_1000rows %>% 
      summarise(x = mean(x),
                y = mean(y),
                z = mean(z))
    })

across_bench %>%
  separate(expression,
           into = c("fun", "groups"),
           convert = TRUE) %>% 
  ggplot(aes(x = groups, y = median, group = fun, color = fun)) +
  geom_line() + 
  scale_y_bench_time(base = NULL)

```
