---
title: "Why dplyover?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why dplyover?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{tidyr}
  %\VignetteDepends{bench}
  %\VignetteDepends{ggplot2}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(
  tibble.print_min = 4,
  tibble.max_extra_cols = 8,
  digits = 2,
  crayon.enabled = FALSE,
  cli.unicode = FALSE
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(dplyover)
library(ggplot2)
library(bench)
```

## Introduction

{dplyover} extends {dplyr}'s functionality by building a function family
around `dplyr::across()`.

The goal of this *over-across function family* is to provide a concise and
uniform syntax which can be used to create columns by applying functions to
vectors and/or sets of columns in {dplyr}. Ideally, this will:

- **reduce the amount of code** to create variables derived from existing colums, 
which is especially helpful when doing explanatory data analysis (e.g. lagging, 
collapsing, recoding etc. many variables in a similar way).
- **provide a clean {dplyr} approach** to create many variables which are
calculated based on two or more variables.
- **improve our mental model** so that it is easier to tackle problems where the
solution is based on creating new columns.

To illustrate the points above it is helpful to look at a couple of use cases.
This vignette will (i) look at the four examples, (ii) present some alternatives
that do not rely on {dplyover} and (iii) elaborate on pros and cons.

## Use cases and workarounds

In this section we will look at three different use cases of {dplyover}:

1. Creating several lagged variables for a set of columns
1. Applying functions to a set of variable pairs
1. Summarizing data in wide format


#### Creating several lagged variables for a set of columns

In exploratory data analysis of time series data, creating a range of lagged
variants of several variables is not an uncommon use case. Let's take the
`diamonds` data set from {ggplot2} and create five lagged versions of columns
'x', 'y' and 'z' by positions `1` to `5`. `crossover` allows us to iterate a set
of columns over a vector:
```{r}
# crossover
diamonds %>% 
  mutate(crossover(c(x,y,z),
                   1:5,
                   list(lag = ~ lag(.x, .y)),
                   .names = "{xcol}_{fn}{y}"))
```
While this was easy, there are of course other ways to approach this problem. We
could use `dplyr::across` and just insert the call to `lag` five times manually
in the `.fns` argument:
```{r, eval = FALSE}
# across
diamonds %>% 
  mutate(across(c(x,y,z),
                list(lag1 = ~ lag(.x, 1),
                     lag2 = ~ lag(.x, 2),
                     lag3 = ~ lag(.x, 3),
                     lag4 = ~ lag(.x, 4),
                     lag5 = ~ lag(.x, 5))))
```
However, with an increasing size of lagged variables or functions^#^ this
approach becomes inefficient in terms of repeating similar code patterns (^#^ for
example if we'd also wanted to create five `lead` variables per column).

Another option is using {purrr}'s `map_dfc` inside `dplyr::mutate`:
```{r, eval = FALSE}
# across and map_dfc
diamonds %>% 
  mutate(across(c(x,y,z),
                ~ map_dfc(set_names(1:5, paste0("lag", 1:5)),
                          function(y) lag(.x, y))
                )) %>% 
  do.call(data.frame, .)
```
This approach is pretty close to `crossover`, but we need to take care of two 
things. First, we have to create the variable names using `purrr::set_names`,
otherwise our new variables would be named `x...1`, `x...2` and so on. Second,
the original output of calling `purrr::map_dfc` within `dplyr::across` on three
columns are three `tibble`s, each containing the lagged variants of one column.
To turn each `tibble` into five proper columns of our original `diamonds` data
we need to pipe the final result into a `do.call(data.frame, .)`. All in all,
although staying in the tidyverse, the `map` call nested in `across`, together
with `set_names` and the final `do.call` make this approach less readable.

Finally, another approach would be to build a custom function and use
`purrr::reduce2` to call several `dplyr::mutate` calls in a row, each with the
input of the last call's output.
```{r, eval = FALSE}
# custom function with purrr::reduce
create_lags <- function(df, .x, .y) {
    mutate(df, "{.x}_lag{.y}" := lag(!! sym(.x), .y))
}

diamonds %>% 
  reduce2(rep(c("x", "y", "z"), 5),
          rep(1:5, 3),
          create_lags,
          .init = .)
```
This approach is very clever, and {purrr}'s `reduce` function is a pipe-friendly
approach which works great with {dplyr}'s one-table verbs. However, here too we
have to take care of two things: First, using NSE in the custom function requires
some advanced knowledge of the tidyverse. (i) How do we create nice names on the 
lefthand side of the walrus operator `:=`? (ii) How do we evaluate strings as
column names? Second, figuring out how to setup `reduce2` to work with our custom
function. Our original data goes in to `.init` argument, the arguments we want
to loop over need to be repeated in a way that ensures that each combination of
elements is created.

Finally, we compare the performance of each operation. The table below shows that
using `dplyr::across` with repititive code parts is the most performant approach, 
while using a custom function with `reduce` is the least performant. Using
`map_dfc` within `across` does not perform much worse than `across` alone and
also `crossover` is not too far off in terms of speed. The big gain of using
`crossover` is that it either (i) saves us from typing repetitive code (compared
to `dplyr::across`) or it reduces the cognitive load of figuring out how to use 
advanced tidyverse workarounds (like `reduce` or a `map` call nested in `across`).

```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
create_lags <- function(df, .x, .y) {
    mutate(df, "{.x}_lag{.y}" := lag(!! sym(.x), .y))
}

crossover_bench <- bench::mark(iterations = 50L, check = FALSE, 
  across = {
    diamonds %>% 
      mutate(across(c(x,y,z),
                    list(lag1 = ~ lag(.x, 1),
                         lag2 = ~ lag(.x, 2),
                         lag3 = ~ lag(.x, 3),
                         lag4 = ~ lag(.x, 4),
                         lag5 = ~ lag(.x, 5))))
  },
  across_map_dfc = {
    diamonds %>% 
      mutate(across(c(x,y,z),
                    ~ map_dfc(set_names(1:5, paste0("lag", 1:5)),
                              function(y) lag(.x, y)))) %>% 
    do.call(data.frame, .)
  },
  
  custom_fct_reduce = {
    diamonds %>% 
  reduce2(rep(c("x", "y", "z"), 5),
          rep(1:5, 3),
          create_lags,
          .init = .)
  },
  crossover = {
    diamonds %>% 
      mutate(crossover(c(x,y,z),
                       1:5,
                       list(lag = ~ lag(.x, .y)),
                       .names = "{xcol}_{fn}{y}"))
  }
)
crossover_bench[c("expression", "median", "mem_alloc")]

```

#### Applying functions to a set of variable pairs

Some data sets contain many variable with similar patterns. A 'tiny' example is
the `ìris` data set, containing columns with combinations of 'Sepal' and 'Petal'
with 'Length' and 'Width'. In the example below we want to calculate the product
and the sum of both pairs of 'Length' and 'Width' variables. `across2` 
works just like `dplyr::across`, but let's us choose *two* sets of columns -
`.xcols` and `.ycols`. `across2` allows us to use a special glue specification,
`{pre}`, in its `.names` argument, which will evaluate to the common prefix of
each variable pair. 

```{r}
iris %>%
  mutate(across2(ends_with("Length"),
                 ends_with("Width"),
                 .fns = list(product = ~ .x * .y,
                             sum = ~ .x + .y),
                 .names = "{pre}_{fn}")) %>%
  glimpse
```

While in this case a simple call to `mutate` (defining two new varialbes) would
have done the job, `across2` really comes in handy when we are dealing with  
data sets with many variables that need to be computed in a similar way.

Again, there are different workarounds to produce the same result. One option is
to use `dplyr::across` and `dplyr::cur_column`. First we shorten the names of
'Sepal.Length' and 'Petal.Length' to their stem using `dplyr::rename_with`.
Then we can construct the names of the 'Width' variables using `paste0`. Below we
use `get` (alternatively we could also use `eval(sym())`) on the so constructed
names to get the underlying values. While this comes close to our desired result,
we probably want to restore the original names. Unfortunatly, this requires an
additional setup, which is why we rename the variables manually below.

```{r, eval = FALSE}
iris %>%
  rename_with(.fn = ~ gsub(".Length", "", .x),
              .cols = ends_with("Length")) %>% 
  mutate(across(c(Sepal, Petal), 
                list(product = ~ .x * get(paste0(cur_column(), ".Width")),
                     sum = ~ .x + get(paste0(cur_column(), ".Width"))))) %>% 
  rename(Sepal.Length = Sepal,
         Petal.Length = Petal)
```

Another approach is again {purrr}'s `reduce2` in combination with a custom 
function. Apart from the data, `df`, our custom function takes two strings as 
arguments: `.x` is the common stem of our variable pairs, and `.y` which is the
name of the operation we want to perform. The downside of this approach was
already mentioned above: (i) advanced NSE knowledge and (ii) figuring out how 
to use `reduce2`, especially building the inputs using `rep`.

```{r, eval = FALSE}
create_vars <- function(df, .x, .y) {
    mutate(df, "{.x}_{.y}" := switch(.y,
                                     sum = !! sym(paste0(.x, ".Width")) +
                                           !! sym(paste0(.x, ".Length")),
                                     product = !! sym(paste0(.x, ".Width")) *
                                               !! sym(paste0(.x, ".Length")))
    )
}

iris %>% 
  reduce2(rep(c("Sepal", "Petal"), each = 2),
          rep(c("product", "sum"), 2),
          create_vars,
          .init = .) 
```

Comparing the performance of the three operations above shows that
`dplyr::across` is, as always, the most performant. This time, `reduce2` follows
closely and `across2` is the least performant. Although showing a lower 
performance, `across2` has several advantages: 
1. No knowledge of NSE or {purrr}'s `reduce` is needed.
1. Its the most readable approach.
1. It allows us to produce nice names, including a common pre- or suffix.
1. It does not depend on the variables having to share a common stem. 

```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
create_vars <- function(df, .x, .y) {
  mutate(df, "{.x}_{.y}" := switch(.y,
                                   sum = !! sym(paste0(.x, ".Width")) + !! sym(paste0(.x, ".Length")),
                                   product = !! sym(paste0(.x, ".Width")) * !! sym(paste0(.x, ".Length")))
  )
}

across2_bench <- bench::mark(iterations = 50L, check = FALSE, 
  across = {
    iris %>%
      rename_with(.fn = ~ gsub(".Length", "", .x),
                  .cols = ends_with("Length")) %>% 
      mutate(across(c(Sepal, Petal), 
                    list(product = ~ .x * get(paste0(cur_column(), ".Width")),
                         sum = ~ .x + get(paste0(cur_column(), ".Width"))))) %>% 
      rename(Sepal.Length = Sepal,
             Petal.Length = Petal)
  },
  custom_fct_reduce = {
    iris %>% 
      reduce2(rep(c("Sepal", "Petal"), each = 2),
              rep(c("product", "sum"), 2),
              create_vars,
              .init = .) 
  },
  across2 = {
    iris %>%
      mutate(across2(ends_with("Length"),
                        ends_with("Width"),
                        .fns = list(product = ~ .x * .y,
                                    sum = ~ .x + .y),
                       .names = "{pre}_{fn}"))
  }
)
across2_bench[c("expression", "median", "mem_alloc")]

```

#### Create columns derived from a comma separated character vector

{dplyover} helps us to think in terms of columns that we want to create. A good
example, when this kind of thinking is useful, is when dealing with a comma 
separated character vector. In the example below the column 'csat_open' contains
one or several comma separated reasons why a good or bad satisfaction rating was
given. We want to transform the data so that each unique reason in 'csat_open'
gets its own column in form of a dummy variable: taking `1` when the reason was
given and `0` otherwise. To do this we use `over` together with `dist_values`
which is a wrapper around `unique` that splits comma separated strings when a
separator is provided in `.sep`:

```{r, eval = TRUE}
csat %>%
  transmute(over(dist_values(csat_open, .sep = ", "),
              ~ as.integer(grepl(.x, csat_open)),
              .names = "rsp_{x}",
              .names_fn = ~ gsub(" ", "_", .x))) %>%
  glimpse # for better printing
```

`over` makes it easy to think about the problem we are dealing with. We just
iterate over all unique values in `csat_open` and check with `grepl` if the 
value is present. The first two arguments would suffice to produce a useable 
result. Here, we additionally use the `.names` and `.names_fn` arguments to
create nice column names by appending a prefix ('rsp_') and replacing the white
spaces with underscores.

A tidyverse approach that comes close to `over`s functionality is to use
`purrr::map_dfc` inside `mutate`. The only difference is that we have to create
a named vector, which holds the new column names as names, and the unique
values of "csat_open" as values. This makes this approach less pipe-friendly.
```{r, eval = FALSE}
var_nms <- strsplit(csat$csat_open, ", ") %>% unlist %>% unique %>%
  set_names(., paste0("rsp_", gsub(" ", "_", .)))

csat %>%
  transmute(map_dfc(var_nms,
              ~ as.integer(grepl(.x, csat_open))))
```

While `purrr::map_dfc` is a valid workaround, it is less known and therefore not
often used in the wild (at least this is my anecdotal evidence that comes from
looking at answers to similar questions on Stackoverflow). 

A more common way to tackle this problem within the tidyverse would be to use
{tidyr}. Here we would first use `separate_rows` and make the data longer.
Then we'd use `pivot_wider` to generate the columns we are interested in. 
Although the code is not much longer, we have to do some thinking about how to
set up `pviot_wider` to yield our desired result. And although we make use of a
lot of `pivot_wider`s arguments, the column names still require a finishing
touch.

```{r, eval = TRUE}
csat %>%
  separate_rows(csat_open, sep = ", ") %>% 
  pivot_wider(names_from = csat_open,
              names_prefix = "rsp_",
              names_repair = "universal",
              values_from = csat_open,
              values_fn = function(x) 1,
              values_fill = 0) %>%
  select(starts_with("rsp_")) %>% # for better printing
  glimpse # for better printing
```

Another approach is to (i) get all unique values in "csat_open", (ii) create 
empty columns with zeros, and then (iii) use `cur_column` with `grepl` inside
`dplyr::across` to test if current column name is present in "csat_open". This 
approach has two negative aspects. First, it is not pipe-friendly. Second, 
creating nice column names isn't easy. This comes as a surprise, since
`dplyr::across` has a `.names` argument. This lets us append a prefix, but it 
does not let us replace whitespaces with underscores. Further, once we make use
of the `.names` argument, new copies of the input columns are generated, so we
need to get rid of the old columns.

```{r, eval = TRUE}
var_nms <- strsplit(csat$csat_open, ", ") %>% unlist %>% unique

# instead of the set_names below we could do the following in base R:
# csat[, var_nms] <- 0
csat %>% 
  transmute(!!! set_names(rep(0, length(var_nms)), var_nms), 
            across(all_of(var_nms), 
                   ~ grepl(cur_column(), csat_open) %>% as.integer,
                   .names = "rsp_{col}")) %>%
  select(!all_of(var_nms)) %>% 
  glimpse # for better printing
```

If we look at the performance of each approach, we can see that `map_dfc` and
`across` are fastest, while `pivot_wider` is the slowest. `over` ranges 
somewhere in between.

```{r, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
csat2 <- csat
over2_bench <- bench::mark(iterations = 50L, check = FALSE, 
  map_dfc = {
    local({var_nms <- strsplit(csat$csat_open, ", ") %>% unlist %>% unique %>%
      set_names(., paste0("rsp_", gsub(" ", "_", .)))

    csat %>%
      transmute(map_dfc(var_nms,
                  ~ as.integer(grepl(.x, csat_open))))
  })},
  pivot_wider = {
    csat %>%
      separate_rows(csat_open, sep = ", ") %>% 
      pivot_wider(names_from = csat_open,
                  names_prefix = "rsp_",
                  names_repair = "universal",
                  values_from = csat_open,
                  values_fn = function(x) 1,
                  values_fill = 0)
  },
  across = {
   local({var_nms2 <- strsplit(csat$csat_open, ", ") %>% unlist %>% unique;
   csat2[, var_nms2] <- 0
    csat2 %>%
      transmute(across(all_of(var_nms2),
                       ~ grepl(cur_column(), csat_open) %>% as.integer,
                       .names = "rsp_{col}")) 
   # %>%
   #    select(!all_of(var_nms))
  })},
  over = {
    csat %>%
      transmute(over(dist_values(csat_open, .sep = ", "),
                  ~ as.integer(grepl(.x, csat_open)),
                  .names = "rsp_{x}",
                  .names_fn = ~ gsub(" ", "_", .x)))
  }
)
over2_bench[c("expression", "median", "mem_alloc")]

```


#### Summarizing data in wide format
Sometimes we want to summarize data in wide format. In the example below we are
looking at the `csat` data from a customer experience survey. We would like 
to compare the customers of each "type" by looking at how many percent give
which customer satifaction rating (column "csat"). If we want each category of
"csat" in a new column, then `over` makes this pretty easy:

```{r, eval = TRUE}
csat %>%
  group_by(type) %>%
  summarise(over(levels(csat),
                 ~ mean(csat == .x)))
```
Of course there are many cases, where we do want the output to be in long format.
For example when creating a graph with {ggplot2}. The summary in wide format, 
however, is useful when we want to print a table in a report, for example.

As already seen above, we can {purrr}'s `map_dfc` inside `summarise` to create
an identical output:
```{r, eval = FALSE}
csat %>%
  group_by(type) %>%
  summarise(map_dfc(set_names(levels(csat), levels(csat)),
                    ~ mean(csat == .x)))
```
Code-wise this comes pretty close to `over`. The major drawback is, again, that
we have to take care of the column names with `set_names`.

Finally, we can create the same output using regular {dplyr} and {tidyr} 
functions:
```{r, eval = FALSE}
csat %>%
  group_by(type, csat) %>%
  summarise(n = n()) %>% 
  mutate(n = prop.table(n)) %>%
  pivot_wider(values_from = n,
              names_from = csat)
```
The first part (up until `pivot_wider`) would be our go-to approach if the result
should be in long format. Compared to `over` the `pivot_wider` approach is not
only more verbose, it also has a higher cognitive load. 

Looking at the performance of each approach, we can see that `map_dfc` is
fastest, followed by `over`, which is a little slower, while `pivot_wider` is
the least performant.
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
over_bench <- bench::mark(iterations = 50L, check = FALSE, 
  over = {
    csat %>%
  group_by(type) %>%
  summarise(over(dist_values(csat),
                 ~ mean(csat == .x)))
  },
  pvior_wider = {
    csat %>%
  group_by(type, csat) %>%
  summarise(n = n()) %>% 
  mutate(n = prop.table(n)) %>%
  pivot_wider(values_from = n,
              names_from = csat)
  },
  map_dfc = {
    csat %>%
  group_by(type) %>%
  summarise(map_dfc(set_names(levels(csat), levels(csat)),
                    ~ mean(csat == .x)))
  }
)
over_bench
```



Summing up ... .


